---
title: "A database for your data-files with duckdb"
author: "Thomas Brand"
date: "2026-07-05"
categories: [news, code, arrow, duckdb]
draft: true
---

```{r}
#| label: setup
#| include: FALSE

# delete duckdb database
fs::file_delete("./myDuck.db")
#here::i_am("posts/database-for-your-files-duckdb") 
```

# Abstract

# Motivation

## Setup

During your work on projects or in the data science department of a corporation you collect a lot of data form several data sources, different applications, different data bases etc.

In order to achieve your goals you will have to combine informations from several data sources, e.g. join tables, add informations and so on.

I assume it takes too long to load these data from the different sources in raw form and transform them in order to be able to work with them. So what you have done is:

- you downloaded your data

- you transformed your data in the way that suits your needs (e.g. convert all dates to the same format, filled in missing values, omitted rows that aren't necessary for your work and so on)

- you saved your data on disk preferably in a data format that is made for data analysis (e.g. the parquet-format)

Sometimes you save your data in files with suffixes for month and years.

Maybe you have made yourself a special table with informations that aren't in any tables of your applications, but are none the less useful/needed for your analysis or presentation of the data. Think of a classification-information of your products where for analysis A it has to be in category 1 and for analysis B it has to be in category 2.

Over times you accumulated a lot of files in different locations on disk and with some luck you can still tell where to find the data files.

## What probably happened

::: callout-note
At least, this is what I did.
:::

You have built loading functions for each type of file that you use.

If it is just one big data file or several parquet files in a directory or with partitioning you can just hard code the path of the data into the loading function.

```{r}
#| label: loading function
#| eval: FALSE

# loading function with column selection
loadTableA = function(path = "./dir1/datafile.parquet",
                      columns = NULL) {
  arrow::read_parquet(file = path,
                      col_select = columns)
}
```

::: callout-note
If you use the open_dataset-function of the arrow-Package (see <https://arrow.apache.org/docs/r/reference/open_dataset.html>) you can even combine row filtering features with your loading-function.

For a detailed explanation of what you can do with the[{arrow}](https://arrow.apache.org/docs/r/index.html) package please see

Richardson N, Cook I, Crane N, Dunnington D, Fran\<U+00E7\>ois R, Keane J, Mecum B, Moldovan-Gr\<U+00FC\>nfeld D, Ooms J, Wujciak-Jens J, Apache Arrow (2026). *arrow: Integration to 'Apache' 'Arrow'*. R package version 24.0.0, <https://github.com/apache/arrow/>.
:::

If you saved your files in different locations depending on certain conditions e.g. the month and year you generated the data you probably combined the solution above with a file finding function.

```{r}
#| label: file finding and loading
#| eval: FALSE

# function to find a file
findFileTableA = function(month = month(Sys.Date()),
                          year = year(Sys.Date())) {
  return(
    normalizePath(
      file.path(".",
                "dir1",
                year,
                month),
      mustWork = TRUE)
  )
}

# loading function with column selection
loadTableA = function(path = NULL,
                      columns = NULL) {
  arrow::read_parquet(file = path,
                      col_select = columns)
}

# using it
pathToFile = findFileTableA(5,2026)
tab = loadTableA(pathToFile,
                 c("col1","col5"))

```

::: callout-note
I used pretty much base R functions. Of course you can use more sophisticated file/path function e.g. in the [{fs}](https://fs.r-lib.org) package
:::

The afore mentioned files with additional informations are probably stored in your packages that you have built in order to be always on hand for your analysis.

::: callout-caution
This strategy works!

It works for you, for your team, for R.

But it gets difficult if you want to share all your data with another department. They will have to learn your conventions, use your code.
:::

# Solution - duckdb

## What we strive for

We would like to have

- one single entry point for all your data

- an entry point, that can be broadly used

- an entry point that will not have to be regenerated if you update your data.

In short: we are talking about putting all your data in a database. But you will have to change all your data-connections as well.

Copying your data to a database will double your disk space needed for data.

And don't underestimate the amount of time and knowledge it requires to administrate a database.

::: callout-note
[Duckdb](https://duckdb.org) and the [{duckdb}](https://r.duckdb.org) package offers you a solution for this kind of problem.

For a detailed explanation of what you can do with duckdb please see

Mühleisen H, Raasveldt M (2026). *duckdb: DBI Package for the DuckDB Database Management System*. R package version 1.5.3, <https://r.duckdb.org/>.
:::

## Prepare some data

For our demonstration we prepare some data. For this we will use the {nycflights23} data.

```{r}
#| label: data preparation parquet
#| messages: FALSE

library(nycflights23)
library(arrow)

write_dataset(dataset = flights,
              path = here::here("database/flights"),
              format = "parquet",
              partitioning = "dest")

write_parquet(x = airlines,
              sink = here::here("database/airlines.parquet"))

write_parquet(x = airports,
              sink = here::here("database/airports.parquet"))
write_parquet(x = planes,
              sink = here::here("database/planes.parquet"))

write_parquet(x = weather,
              sink = here::here("database/weather.parquet"))

```

With this we have serveral parquet-files with tables and even on directory with segmented data.

## Setup a database

For this we create a duckdb on disk an fill it with views to the parquet-files on disk

```{r}
#| label: creating duckdb 

library(duckdb)

# create a database driver
drv = duckdb(dbdir = "./myDuck.db")

# get a connection to the database
con = dbConnect(drv)

# create View for the dataset
dbExecute(con,
          "CREATE VIEW flights AS
                SELECT * from read_parquet('database/flights/')")

# show tables
dbListTables(con)
```

Query this new table

```{r}
#| label: query the database

result = dbGetQuery(conn = con,"
                    SELECT tailnum, dep_delay, arr_delay 
                    FROM flights
                    WHERE tailnum = 'N807JB' and day = 1
                    ")

result
```

Create Views for every parquet-file in a directory and/or replace it if it already exists

```{r}
#| label: dynamik view creation

parquetFiles = list.files("database", 
                          pattern = "*.parquet")

library(glue)

for (f in parquetFiles) {
  fwoExt = tools::file_path_sans_ext(f)
  fwithPath = file.path("database", f)
  sqlString = glue_sql("
                       CREATE OR REPLACE VIEW
                       {`fwoExt`} AS
                       SELECT * from read_parquet({`fwithPath`})", 
                       .con = con)
  
  
  dbExecute(con,
            sqlString)
}

dbListTables(con)
```

Query the airports

```{r}
#| label: query airports 1

airp = dbGetQuery(conn = con, "
                  SELECT * 
                  FROM airports
                  LIMIT 5")
airp
```

Query faa = "BMX"

```{r}
dbGetQuery(conn = con,"
           SELECT *
           FROM airports
           WHERE faa = 'BMX'")
```

Don't forget to close your database connection

```{r}
#| label: close database

dbDisconnect(con)
```

## Size of data base

Let's look at the size of the database

```{r}
#| label: size of database

fs::file_size("./myDuck.db")
```

The data itself are quite larger

```{r}
#| label: size of data

fs::dir_ls("./database", recurse = TRUE) |>
  fs::file_size() |>
  sum()
```

## Change or append datasets

We'll change an append die airports data

```{r}
#| label: manipulate airport data

library(data.table)
airports = setDT(read_parquet("./database/airports.parquet"))
bmx = airports[faa == "BMX"][, name := "Fancy Airport Name"]
airports[faa == "BMX", name := paste("Very",name, sep = " ")]
airports = rbind(airports, bmx)
write_parquet(airports, sink = "./database/airports.parquet")
```

Now we open the data base again and repeat the query for faa = "BMX

```{r}
#| label: second query

# create a database driver
drv = duckdb(dbdir = "./myDuck.db")

# get a connection to the database
con = dbConnect(drv)

# query
result = dbGetQuery(conn = con,"
                    SELECT *
                    FROM airports
                    WHERE faa = 'BMX'")

dbDisconnect(con)

result
```

We can see, that it is really a VIEW we have here, so that a change in the underlying data reflects to the data base.

# Closing Remarks

The code above to create a VIEW from every parquet-file in a directory is in my opinion quite useful if you have your classification-information tables or mapping tables inside your package. These files are stored in the data-directory of your package in a csv or Rdata format.

While duckdb comes with a function to read csv-files (see <https://duckdb.org/docs/current/data/csv/overview> ) I don't recommend it as csv-files are easy to read for the human eye but are difficult to interpret for the machine. Think of the different way you can write a date.

So for this special purpose I recommend that you convert your Rdata files into parquet-format. This is quite easily done by inserting an R-script at the end of the R-directory of your package, that converts every Rdata-file or Rds-file of the data-directory of your package into parquet-format. You will end up with two formats of the same data in your data-directory - one Rdata and one parquet of each data-file.
